Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-06-08 ICML/ManuscriptWithCommentsJP.Rmd
Comments on thesis manuscript

@@

A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice]. With MICE, imputed values are drawn from the posterior predictive distribution of the missing values. The algorithm is named after its iterative nature: a multivariate distribution is obtained by iterating over a sequence of univariate imputations. Iteration, however, also introduces a potential threat to the validity of the imputations: What if the algorithm has not converged? Are the imputations then to be trusted? And can we rely on the inference obtained on the completed data?

These remain open questions since the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18]. There is no scientific consensus on how to evaluate the convergence of imputation algorithms [@zhu15; @taka17]. Moreover, the behavior of such algorithms under certain default imputation models (e.g., 'predictive mean matching') is an entirely open question [@murr18]. Therefore, algorithmic convergence should be monitored carefully---although this is not straightforward. Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence [@gelm13]. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostic methods may only identify signs of *non*-convergence [@hoff09]. Several non-convergence diagnostics exist, but it is not known whether these are appropriate within the imputation framework. <!-- "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->

@@

Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection. After running the imputation algorithm for a certain number of iterations, researchers are encouraged to produce traceplots. In a traceplot, a scalar summary of the state-space of the algorithm $\theta$ is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}. The default scalar summaries to inspect for the MICE algorithm are chain means and chain variances. Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends [@buur18, $\S$ 6.5.2]. 
<!-- Recommendation @buur18: "the different streams should be freely intermingled with one another, without showing any definite trends". -->

@@

We propose $\lambda_{1}$ as such a scalar summary. We define $\lambda_{1}$ as the first eigenvalue of the variance-covariance matrix in the $m$ completed datasets. The eigenvalues of a variance-covariance matrix $S$ summarize the total set of covariances in the data. Let $\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ... \geq \lambda_{j, \ell}$ be the eigenvalues of $S_\ell$ in each imputation. The first eigenvalue, $\lambda_{1,\ell}$, then summarizes the largest possible amount of covariance in each completed dataset {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}. By definition, $\lambda_{1}$-values are equal to the variance of the first component that would be obtained by performing principal component analysis (PCA) on the completed data. As scalar summary, $\lambda_{1}$ has the appealing property that it is not dependent on the model of scientific interest, yet still summarizes the multivariate state-space of the algorithm.

@@

There are many diagnostic tools to identify non-convergence in iterative (MCMC) algorithms [@broo98; @elad06]. We consider only two of them that may be appropriate for imputation algorithms---one to monitor signs of non-mixing, and one for non-stationarity. As recommended by e.g. @cowl96 we will use the potential scale reduction factor $\widehat{R}$ to evaluate mixing ['Gelman-Rubin statistic'; @gelm92], and autocorrelation to diagnose trending [$AC$; @scha97; @gelm13]. With a recently proposed adaptation, $\widehat{R}$ might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated [@veht19]. Therefore, we will evaluate the appropriateness of both $\widehat{R}$ versions, in addition to $AC$. Other methods are outside the scope of this study because they e.g. assume that values within chains represent independent samples, whereas the MICE algorithm only uses the final iteration to produce imputations.

@@

```
# pseudo-code of simulation 
1. simulate data 
for (number of simulation runs from 1 to 1000)
 for (proportion of missing cases between 5% and 95%)
  2. create missingness
  for (number of iterations from 1 to 100)
   3. impute missingness
   4. perform analysis of scientific interest
   5. compute non-convergence diagnostics 
   6. pool results across imputations
   7. compute performance measures
 8. combine outcomes of all conditions
9. aggregate outcomes across simulation runs

@@

We evaluate the diagnostic methods to identify non-convergence against these performance measures. $\widehat{R}$ and $AC$ should identify non-convergence in simulation conditions, where $\bar{Q}$ is *not* an unbiased, confidence-valid estimate of $Q$.

@@

```{r Qs, fig.height = 6.8, fig.cap="Impact of non-convergence on statistical inferences. Depicted are the bias, coverage rate (CR) and confidence interval width (CIW) of the estimates $\\bar{Q}$. We present the subset of $Q$s with the worst performance in terms of bias: the mean and standard deviation of $Y$, and the regression coefficient of $X_1$. The solid gray line depicts the targets (unbiased estimates and nominal coverage rate)."}
mean_bias + sd_bias + Rsq_bias + est_bias + est_cov + est_ciw + plot_annotation(tag_levels = "A", tag_suffix = ".") + plot_layout(ncol = 2, guides = "collect") 
```

@@

```{r rhats-acs, fig.height = 6.8, fig.cap="Non-convergence diagnostics $\\widehat{R}$ and $AC$ applied on several $\\theta$s. The left-hand side of the figure contains $\\widehat{R}$-values, and the right-hand side contains $AC$-values. Depicted in the rows are the scalar summaries $\\theta$: chain mean, chain variance, the quantity of scientific interest $\\hat{Q}$, and the first eigenvalue of the variance-covariance matrix $\\lambda_1$. The dashed gray lines depict the different thresholds to diagnose non-convergence."}
mean_Rh + mean_AC + var_Rh + var_AC + est_Rh + est_AC + PCA_Rh + PCA_AC + plot_layout(guides = "collect", ncol = 2) + plot_annotation(tag_levels = "A", tag_suffix = ".")
```

### $\theta=$ chain mean. 
As expected, $\widehat{R}$ lowers with the number of iterations. After an initial 'dip' in conditions where $T=3$, we observe a gradual decrease in $\widehat{R}$-values with every additional iteration. This implies improving convergence, until the decrease tapers off after 30 to 40 iterations. Using the threshold $\widehat{R}>1.2$, we diagnose non-convergence in conditions where $T\leq7$, whereas the threshold 1.1 is exceeded when $T\leq13$. The very strict threshold 1.01 is surpassed in all conditions $T\leq100$ (not shown), and therefore not informative under the current specifications. With this set of methods, we fail to identify non-convergence in conditions with persistent bias due to missingness.

@@

### $\theta=\hat{Q}$.
The $\widehat{R}$-values in Figure \ref{fig:rhats-acs}E show a similar trend across iterations as we observed for chain means and chain variances. Only conditions with a very high  proportion of missing cases ($p_{\rm inc}\geq.75$) diverge from the earlier observations. The $\widehat{R}$-values in these conditions do not taper off after 30 to 40 iterations, but rather between 40 to 50 iterations. The highest number of iterations at which non-convergence is still diagnosed according to the thresholds are 11 for $\widehat{R}>1.2$, 23 for $\widehat{R}>1.1$, and 100 for $\widehat{R}>1.01$ (not shown).

@@

$AC$-values are systematically higher for this $\theta$ than for other scalar summaries. The $AC$-values also taper off at a later point in the iterations. This suggests improving stationarity up-to ten, or even thirty iterations, depending on the missingness condition. According to the threshold, we should diagnose non-convergence at some point in any missingness condition. However, this occurs only after reasonably decreasing $AC$-values are obtained---again suggesting that non-convergence is incorrectly diagnosed.

@@

3. We hypothesized that the recommended thresholds to diagnose non-convergence with $\widehat{R}$ would be too stringent for iterative imputation applications, compared to the number of iterations necessary for statistically valid inferences. This hypothesis is partially supported: On the one hand, the 1.01 and 1.1 thresholds indeed seem to over-estimate the signs of non-convergence. On the other hand, however, we observed that the traditional threshold of 1.2 roughly corresponds to the number of iterations required to obtain approximately unbiased, confidence-valid estimates.

@@

5. Our final hypothesis was that multivariate $\theta$s would be better at detecting non-convergence than univariate $\theta$s. Multivariate $\theta$s did indeed show superior performance under certain conditions (e.g., multivariate $Q$s and extreme missingness).

>>>>

**Comments Gerko: "The thesis by Hanne covers algorithmic convergence in iterative imputation; a topic that has not received much attention but is ever so important as most imputation software packages draw inference from iterative imputation procedures. The validity of this inference all depends on the state space of the algorithm at the final iteration. Hanne examined the relevant streams of literature; identified the specific components and substantive model dependent and substantive model-independent evaluation measures that are necessary to identify non-convergence and assess the validity of obtained inferences and was able to simulate and investigate the plausibility of identifiers for non-convergence for iterative imputation algorithms. Her work results in a guide- line for practice. ... There is clear problem, existing methodology, a new proposed theta for identifying non-convergence, and the simulation approach results in a clear investigation into the convergence properties of iterative imputation algorithms. Hanne proposes guidelines for practice and identifies pitfalls that arise with contemporary evaluation functions and details how to avoid such problems. Her work results in a thorough insight into the convergence properties of iterative imputation algorithms."**

@@

A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice]. With MICE, imputed values are drawn from the posterior predictive distribution of the missing values. The algorithm is named after its iterative nature: a multivariate distribution is obtained by iterating over a sequence of univariate imputations. Iteration, however, also introduces a potential threat to the validity of the imputations: What if the algorithm has not converged? Are the imputations then to be trusted? And can we rely on the inference obtained on the completed data? **Comment jp: "it is not stressed that convergence issues apply to/stem from the Markov chain(s) that is produced by the algorithm."**

These remain open questions since the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18]. There is no scientific consensus on how to evaluate the convergence of imputation algorithms [@zhu15; @taka17]. Moreover, the behavior of such algorithms under certain default imputation models (e.g., 'predictive mean matching') is an entirely open question [@murr18]. Therefore, algorithmic convergence should be monitored carefully---although this is not straightforward. Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence [@gelm13]. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostic methods may only identify signs of *non*-convergence [@hoff09]. Several non-convergence diagnostics exist, but it is not known whether these are appropriate within the imputation framework. **Comment JP: difference from Schafer 2002: "From a theoretical standpoint, this technique is problematic, because the sequence of regression models might not be consistent with a true joint distribution. Technically speaking, these iterative algorithms may never “converge” because the joint distribution to which they may converge does not exist. Nevertheless, simulation work (Brand, 1999) suggests that in some practical applications the method can indeed work well despite the theoretical problems" (p. 169). Also, from Schafer and Olsen (p. 556): "There is no danger in using too large a value of k, because once DA has converged the process remains stationary. On the other hand, ifk is too small, the imputed datasets will not be truly independent and the variability among them may understate the true levels of missing-data uncertainty. Time and computational resources allowing, it is therefore advisable to choose k to be somewhat larger than necessary, to create a margin of safety and ensure that the multiple imputations are truly independent." **

<!-- https://www-tandfonline-com.proxy.library.uu.nl/doi/pdf/10.1207/s15327906mbr3304_5?needAccess=true -->

<!-- "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->

@@

Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection. **Comment JP: not possible to identify visually the point in an MCMC chain where convergence has been reached."** After running the imputation algorithm for a certain number of iterations, researchers are encouraged to produce traceplots. In a traceplot, a scalar summary of the state-space of the algorithm $\theta$ is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}. The default scalar summaries to inspect for the MICE algorithm are chain means and chain variances. Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends [@buur18, $\S$ 6.5.2]. 
<!-- Recommendation @buur18: "the different streams should be freely intermingled with one another, without showing any definite trends". -->

@@

We propose $\lambda_{1}$ as such a scalar summary. We define $\lambda_{1}$ as the first eigenvalue of the variance-covariance matrix in the $m$ completed datasets. The eigenvalues of a variance-covariance matrix $S$ summarize the total set of covariances in the data. Let $\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ... \geq \lambda_{j, \ell}$ be the eigenvalues of $S_\ell$ in each imputation. The first eigenvalue, $\lambda_{1,\ell}$, then summarizes the largest possible amount of covariance in each completed dataset {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}. By definition, $\lambda_{1}$-values are equal to the variance of the first component that would be obtained by performing principal component analysis (PCA) on the completed data. As scalar summary, $\lambda_{1}$ has the appealing property that it is not dependent on the model of scientific interest, yet still summarizes the multivariate state-space of the algorithm.**comment JP: "It is unclear if the lambda_1 is meant for the algorithm in total or is it possible to monitor convergence of single observations or only for the multivariate set of observations?"**

@@

There are many diagnostic tools to identify non-convergence in iterative (MCMC) algorithms [@broo98; @elad06]. We consider only two of them that may be appropriate for imputation algorithms---one to monitor signs of non-mixing, and one for non-stationarity. As recommended by e.g. @cowl96 we will use the potential scale reduction factor $\widehat{R}$ to evaluate mixing ['Gelman-Rubin statistic'; @gelm92], and autocorrelation to diagnose trending [$AC$; @scha97; @gelm13]. With a recently proposed adaptation, $\widehat{R}$ might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated [@veht19]. Therefore, we will evaluate the appropriateness of both $\widehat{R}$ versions, in addition to $AC$. Other methods are outside the scope of this study because they e.g. assume that values within chains represent independent samples, whereas the MICE algorithm only uses the final iteration to produce imputations. **Comment JP: The main point is usually to identify the area where it can be assumed that the chain has converged, it is usually not the question if the chain converges or not."**

@@

```
# pseudo-code of simulation 
1. simulate data 
for (number of simulation runs from 1 to 1000)
 for (missingness proportions 5%, 25%, 50%, 75% and 95%)
  2. create missingness
  for (number of iterations from 1 to 100)
   3. impute missingness
   4. perform analysis of scientific interest
   5. compute non-convergence diagnostics 
   6. pool results across imputations
   7. compute performance measures
 8. combine outcomes of all missingness proportions
9. aggregate outcomes of all simulation runs

@@

We evaluate the diagnostic methods to identify non-convergence against these performance measures. $\widehat{R}$ and $AC$ should identify non-convergence in simulation conditions, where $\bar{Q}$ is *not* an unbiased, confidence-valid estimate of $Q$. **Comment JP: "It is not clear if the object is to do more iterations if the diagnostics has not reached convergence."**

@@

```{r Qs, fig.height = 6.8, fig.cap="Impact of non-convergence on statistical inferences. Depicted are the bias, coverage rate (CR) and confidence interval width (CIW) of the estimates $\\bar{Q}$. We present the subset of $Q$s with the worst performance in terms of bias: the mean and standard deviation of $Y$, and the regression coefficient of $X_1$."}
mean_bias + sd_bias + Rsq_bias + est_bias + est_cov + est_ciw + plot_layout(ncol = 2, guides = "collect") + plot_annotation(tag_levels = "A", tag_suffix = ".")
```

@@

```{r rhats-acs, fig.height = 6.8, fig.cap="Non-convergence diagnostics $\\widehat{R}$ and $AC$ applied on several $\\theta$s. The left-hand side of the figure contains $\\widehat{R}$-values, and the right-hand side contains $AC$-values. Depicted in the rows are the scalar summaries $\\theta$: chain mean, chain variance, the quantity of scientific interest $\\hat{Q}$, and the first eigenvalue of the variance-covariance matrix $\\lambda_1$."}
mean_Rh + mean_AC + var_Rh + var_AC + est_Rh + est_AC + PCA_Rh + PCA_AC + plot_layout(guides = "collect", ncol = 2) + plot_annotation(tag_levels = "A", tag_suffix = ".")
```

### $\theta=$ chain mean. 
As expected, $\widehat{R}$ lowers with the number of iterations. After an initial 'dip' in conditions where $T=3$, we observe a gradual decrease in $\widehat{R}$-values with every additional iteration. This implies improving convergence, until the decrease tapers off after 30 to 40 iterations. Using the threshold $\widehat{R}>1.2$, we diagnose non-convergence in conditions where $T\leq7$, whereas the threshold 1.1 is exceeded when $T\leq13$. The very strict threshold 1.01 is surpassed in all conditions $T\leq100$ (not shown), and therefore not informative under the current specifications. With this set of methods, we fail to identify simulation conditions with persistent bias due to high missingness proportions.

@@

### $\theta=\hat{Q}$.
The $\widehat{R}$-values in Figure \ref{fig:rhats-acs}E show a similar trend across iterations as we observed for chain means and chain variances. Only conditions with very high missingness proportions ($p_{\rm inc}\geq.75$) diverge from the earlier observations. The $\widehat{R}$-values in these conditions do not taper off after 30 to 40 iterations, but rather between 40 to 50 iterations. The highest number of iterations at which non-convergence is still diagnosed according to the thresholds are 11 for $\widehat{R}>1.2$, 23 for $\widehat{R}>1.1$, and 100 for $\widehat{R}>1.01$ (not shown).

@@

$AC$-values are systematically higher for this $\theta$ than for other scalar summaries. The $AC$-values also taper off at a later point in the iterations. This suggests improving stationarity up-to ten, or even thirty iterations, depending on the missingness proportion. Using the threshold, we diagnose non-convergence at some point in all missingness conditions. However, this occurs only after reasonably decreasing $AC$-values are obtained---again suggesting that non-convergence is incorrectly diagnosed.

@@

3. We hypothesized that the recommended thresholds to diagnose non-convergence with $\widehat{R}$ would be too stringent for iterative imputation applications. This hypothesis is only partially supported. $\widehat{R}$ indicated improving convergence until 30 to 40 iterations, while approximately unbiased, confidence-valid estimates were obtained after just seven iterations. This roughly corresponds to the threshold of 1.2. The more stringent threshold 1.1 seems plausible compared to the decrease in $\widehat{R}$-values across iterations. The most recent threshold 1.01 does indeed seem to over-estimate the signs of non-convergence.

@@

5. Our final hypothesis was that multivariate $\theta$s would be better at detecting non-convergence than univariate $\theta$s. Multivariate $\theta$s did indeed show superior performance under certain conditions (e.g., multivariate $Q$s and extreme missingness proportions).
